Reduced plasticity and relative variance in physiological rates of ectotherm populations under climate change

### 2024-06-19 ms.qmd
Dealing with intro comments and R1 acclimation

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining phenotypic variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures driven by the lower proportional increase in physiological rate variance relative to the mean. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that beneficial acclimation responses in some habitats (marine and freshwater) may trade-off with reductions in physiological rate variance that can have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

Phenotypic plasticity is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to proximate environmental conditions making it the first ‘line-of-defense’ against rapid environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses. Acute temperature increases that push organisms from their thermal optimum can be mitigated by plastically adjusting reaction rates in response to chronic temperatures through acclimation [@Seebacher2015-wi; @Havird2020-mj]. Acclimation alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj]. Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures.

@@

Understanding how variability in physiological rates -- traits thought to be closely linked to fitness -- are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change (i.e., the portfolio effect) [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013; @Sanderson2023]. Decreases in phenotypic variance also suggests strong stabilising selection or reflects constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation; @Sanderson2023]. The implications of changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date there are few data testing the importance of variances in this context [@Scheiner2020].

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial habitats are thought to be particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper]. However, this conclusion has been questioned given that marine ectotherms have recently been shown to be closer to their upper thermal limits compared to terrestrial species [@Pinsky2019-fn]. Marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic habitats, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given the observed variation among terrestrial and aquatic ectotherms in their acclimation capacity and the different challenges associated with living in air versus water, the effect of climate warming on variation in physiological traits may differ among terrestrial and aquatic habitats, which may have important cascading effects on energy transfer and productivity between these habitats [@seebacher2023can; @barneche2021warming].

@@

### *Meta-Analysis*
We analysed our data using multilevel meta-analytic (MLMA) and meta-regression (MLMR) models in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms* [vers. `r utils::packageVersion("brms")` @Burkner2017-wg; @Burkner2018-ur; @noauthor_2021-az] and *metafor* [vers. `r utils::packageVersion("metafor")` @Viechtbauer2010-fn].  We fit both Bayesian and frequentist approaches to ensure that our results were consistent, and to create orchard plots more easily [vers. 2.0, @Nakagawa2021-ls; @nakagawaorchard]. In addition, Bayesian methods better protect against type I errors in the presence of complex sources of non-independence [@Nakagawa2021-uf; @Song2021-ba; @Noble2017-mz]. In all cases, frequentist and Bayesian models resulted in the same conclusions. For our Bayesian models, we ran 4 MCMC chains, each with a warm-up of 1000 followed by 4000 sampling iterations keeping every 5 iterations for a total of 3200 samples from the posterior distribution. We used flat Gaussian priors for 'fixed' effects (i.e., $N(0, 10)$) and a student t-distribution for 'random' effects (i.e., $student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing and had converged (i.e., $R_{hat} = 1$). We compared any competing models using Akaike's Information Criteria (AIC) (if frequentist) or Wantabe Information Criteria (WIC) (if Bayesian). We deemed models with the lowest IC value to be best supported if there was a $\Delta IC$ between the competing models of 2 or more. If two models were within 2 $\Delta IC$ units we went with the most parsimonious model.

@@

A phylogeny was derived using the Open Tree of Life (OTL) with the *rotl* package in R (vers. `r utils::packageVersion("rotl")`) [@Michonneau2016-if], and plotted using *ggtree* (vers. `r utils::packageVersion("ggtree")`) [@Yu2017-wj]. We resolved all polytomies in the tree. Any missing taxa were replaced with closely related species and branch lengths were computed using Grafen's method [using power = 0.7, @Grafen1989-pr]. Models fit using correlation matrices computed with different power (p) parameters (from 0.5 – 1.0) has nearly identical AIC. As such, we used an intermediate value of p = 0.7. We used the R packages *ape* (vers. `r utils::packageVersion("ape")`) [@Paradis2019-gx] and *phytools* (vers. `r utils::packageVersion("phytools")`) [@Revell2012-vq] to prune the tree for individual analyses and calculate phylogenetic covariance (or correlation) matrices used in meta-analytic models.

@@

To understand the consequences of human-induced climate change on the potential to impact the opportunity for selection on physiological traits we fit a model that included an interaction between acclimation type, habitat type, latitude and longitude. We assumed that any change in $lnCVR_{Q_{10}}$ across latitude and longitude could vary by habitat type (i.e., an interaction between habitat). We used non-linear tensors for latitude and longitude as any response could be complicated by local factors (e.g., altitude). Our model included random effects of species, trait, phylogeny and study. We predicted the expected change in $lnCVR_{Q_{10}}$ for each wild population in our dataset at the specific populations latitude and longitude. To do this, we first converted the predicted $lnCVR_{Q_{10}}$ to a 1&deg;C change as opoosed to 10&deg;C as follows:

@@

@@

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures driven by the lower proportional increase in physiological rate variance relative to the mean. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that beneficial acclimation responses in some habitats (marine and freshwater) may trade-off with reductions in physiological rate variance that can have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

Reversible phenotypic plasticity, such as physiological acclimation, is relatively rapid and can be fine-tuned to proximate environmental conditions making it the first ‘line-of-defense’ against rapid environmental change [@Dewitt1998; @Scheiner2020]. For example, physiological rates are known to speed up as temperature increases because of the thermodynamic effects on chemical reaction rates -- so called 'acute' temperature responses (@fig-concept). Acute temperature increases that push organisms from their thermal optimum can be mitigated by plastically adjusting reaction rates in response to chronic temperatures through acclimation [@Seebacher2015-wi; @Havird2020-mj]. Acclimation alters acute thermal sensitivity to offset the potentially negative effects of acute temperature changes (e.g., higher energetic demands). Acclimation, however, does not normally result in physiological rates returning to their original state (‘complete’ compensation - *sensu* @huey1999testing), but rather are only partially compensated [@huey1999testing; @Havird2020-mj].

Physiological acclimation is driven by endocrine and epigenetic processes that change the underlying physiology to allow organisms to maintain physiological performance around a fitness optimum despite changes in the environment [@seebacher2019histone; @little2013thyroid; @taff2016endocrine]. However, the focus up to now has been primarily on mean physiological responses neglecting how variability in physiological rates might also be impacted at higher temperatures.

@@

Phenotypic plasticity is expected to evolve in populations when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Thermal variation and predictability are known to differ across major habitats (terrestrial, marine and freshwater), and as such, we expect the extent of plasticity to vary. Species occupying terrestrial habitats often show weak acclimation abilities and so are likely to be particularly vulnerable to climate change given their greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper]. However, this conclusion has been questioned given that marine ectotherms have recently been shown to be closer to their upper thermal limits compared to terrestrial species [@Pinsky2019-fn]. Marine and freshwater habitats appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic habitats, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given the observed variation among terrestrial and aquatic ectotherms in their acclimation capacity and the different challenges associated with living in air versus water, the effect of climate warming on variation in physiological traits may differ among terrestrial and aquatic habitats, which may have important cascading effects on energy transfer and productivity between these habitats [@seebacher2023can; @barneche2021warming].

Understanding how variability in physiological rates are affected by climate change is important because lack of physiological variation can limit responses to selection (i.e., the 'opportunity for selection on a trait') [@Pelletier2012].  Decreases in phenotypic variance also suggests strong stabilising selection or reflects constraints on performance [@Scheiner2020; @Dewitt1998]. Changes in physiological trait variation may also have important ecological consequences by promoting population productivity and stability [@Agashe2009; @kooijman1989population], species coexistence and ecosystem processes [@imura2003genetic; @Bolnick2011; @hart2016variation; @Sanderson2023]. Higher physiological variance in a population may also indicate greater niche breadth which can buffer populations against environmental change (i.e., the portfolio effect) [@Bolnick2011; @Schindler2010; @Zheng2023; @Slatyer2013; @Sanderson2023]. Changes in variance could, therefore, have wide-reaching consequences for understanding the capacity of populations to persist in and adapt to novel environments but to date there are few data testing the importance of variances in this context [@Scheiner2020].

@@

### *Meta-Analysis*
We analysed our data using multilevel meta-analytic (MLMA) and meta-regression (MLMR) models in R (vers. `r paste0(R.Version()$major, ".", R.Version()$minor)`) using *brms* [vers. `r utils::packageVersion("brms")` @Burkner2017-wg; @Burkner2018-ur; @noauthor_2021-az] and *metafor* [vers. `r utils::packageVersion("metafor")` @Viechtbauer2010-fn].  We fit both Bayesian and frequentist approaches to ensure that our results were consistent, and to create orchard plots more easily [vers. 2.0, @Nakagawa2021-ls; @nakagawaorchard]. In addition, Bayesian methods better protect against type I errors in the presence of complex sources of non-independence [@Nakagawa2021-uf; @Song2021-ba; @Noble2017-mz]. In all cases, frequentist and Bayesian models resulted in the same conclusions. For our Bayesian models, we ran 4 MCMC chains, each with a warm-up of 1000 followed by 4000 sampling iterations keeping every 5 iterations for a total of 3200 samples from the posterior distribution. We used flat Gaussian priors for 'fixed' effects (i.e., $N(0, 10)$) and a student t-distribution for 'random' effects (i.e., $student_{t}(3, 0, 10)$). We checked that all MCMC chains were mixing and had converged (i.e., $R_{hat} = 1$). We compared any competing models using Akaike's Information Criteria (AIC) (if frequentist) or Wantabe Information Criteria (WAIC) (if Bayesian). We deemed models with the lowest IC value to be best supported if there was a $\Delta IC$ between the competing models of 2 or more. If two models were within 2 $\Delta IC$ units we went with the most parsimonious model.

@@

A phylogeny was derived using the Open Tree of Life (OTL) with the *rotl* package in R (vers. `r utils::packageVersion("rotl")`) [@Michonneau2016-if], and plotted using *ggtree* (vers. `r utils::packageVersion("ggtree")`) [@Yu2017-wj]. We resolved all polytomies in the tree. Any missing taxa were replaced with closely related species and branch lengths were computed using Grafen's method [using power = 0.7, @Grafen1989-pr]. Models fit using correlation matrices computed with different power (p) parameters (from 0.5 – 1.0) has nearly identical $AIC_{c}$. As such, we used an intermediate value of p = 0.7. We used the R packages *ape* (vers. `r utils::packageVersion("ape")`) [@Paradis2019-gx] and *phytools* (vers. `r utils::packageVersion("phytools")`) [@Revell2012-vq] to prune the tree for individual analyses and calculate phylogenetic covariance (or correlation) matrices used in meta-analytic models.

@@

To understand the consequences of human-induced climate change on the potential to impact the opportunity for selection on physiological traits we fit a model that included an interaction between acclimation type, habitat type, latitude and longitude. We assumed that any change in $lnCVR_{Q_{10}}$ across latitude and longitude could vary by habitat type (i.e., an interaction between habitat). We used non-linear tensors for latitude and longitude as any response could be complicated by local factors (e.g., altitude). Our model included random effects of species, trait, phylogeny and study. We predicted the expected change in $lnCVR_{Q_{10}}$ for each wild population in our dataset at the specific populations latitude and longitude. To do this, we first converted the predicted $lnCVR_{Q_{10}}$ to a 1&deg;C change as opposed to 10&deg;C as follows:

@@

  # Additional model fit as a result of reviewer commenting that acclimation time is assumed to be linear. So, fit a non-linear parameter to see if it improves fit (even though data doesn't suggest it is the case).
   model1.2_nl <- metafor::rma.mv(effect_size ~  max_temp_c + c_accl_time + I(c_accl_time^2) + type + habitat + type:habitat, V = V, 
                              random = list(~1 | record_num,
                                            ~1 | species_full,
                                            ~1 | trait,
                                            ~1 + max_temp_c| spp,
                                            ~1 | obs),
                              R = list(species_full = phylo_cor), dfs = "contain", 
                              test = "t", data = lnRRQ10_data)

    AIC(model1.2_nl, model1.2) # Again, basically the same and go with most parsimonious model

@@

```{r}
  bubble_plot(model1.2, mod = "c_accl_time", group = "spp", ylab = TeX("$lnRR_{Q_{10}}$"), xlab = "Centered Acclimaton Time (days)") + theme_classic() + theme(axis.title  = element_text(size = 30), 
                        axis.text = element_text(size = 26), 
                        axis.text.y = element_text(size = 26),
                        legend.text = element_text(size = 24), 
                        legend.title = element_text(size = 26))
```
